The role of standing variation in geographic convergent adaptation

### 2015-04-02 standing-variation-paper.tex
peter edited through

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage[authoryear]{natbib}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{hyperref}
%\usepackage{latexml} % for \iflatexml
\usepackage{color}
\newcommand{\gc}[1]{{\it\color{blue} #1} }
\newcommand{\mfp}[1]{{\it\color{red} #1} }

@@

\section*{Abstract}
The extent to which populations experiencing shared selective pressures
adapt through a shared genetic response
is relevant to many questions in evolutionary biology. 
In a number of well studied traits and species, 
it appears that convergent evolution within species is common. 
In this paper, 
we explore how standing, deleterious genetic variation contributes to
convergent genetic responses in a geographically spread population,
extending our previous work on the topic.
Geographically limited dispersal slows the spread of each selected allele,
hence allowing other alleles 
-- newly arisen mutants or present as standing variation -- 
to spread before any one comes to dominate the population. 
When such alleles meet, their progress is substantially slowed 
-- if the alleles are selectively equivalent, 
they mix slowly, dividing the species range into a random tessellation,
which can be well understood by analogy to a Poisson process model of crystallization. 
In this framework, 
we derive the geographic scale over which a typical allele is expected to dominate, 
the time it takes the species to adapt as a whole,
and the proportion of adaptive alleles that arise from standing variation.
Finally, we explore how negative pleiotropic effects of alleles before an environment change 
can bias the subset of alleles that get to contribute to a species adaptive response. 
We apply the results to the many geographically localized G6PD
deficiency alleles thought to confer resistance to malaria, \mfp{where the large
mutational target size makes it a likely candidate for adaptation
from standing variation, despite the selective cost of G6PD
deficiency alleles in the absence of malaria.}
We find the numbers and geographic spread of these alleles matches our predictions reasonably well,
\mfp{consistent with the view that they arose from a combination of} from standing variation and new mutations since the advent of malaria.
Our results suggest that much of adaptation may be geographically local 
even when selection pressures are widespread. \mfp{Therefore, we argue
that caution must be exercised when arguing that strongly
geographically restricted alleles are necessarily the outcome of local adaptation.}
We close by discussing the implications of these results for arguments of species coherence and the nature of divergence between species.

@@

Spatial population structure,
as caused for example by geographically limited dispersal, 
also increases the chance of convergent evolution. 
For example, 
\gc{geographically patchy} selection pressures can lead to much higher probability of parallel adaptation than uniform pressures,
since alleles are unable to spread through intervening populations
\citep{RalphCoop:14}. 
In \citet{ralphcoop2010} we formulated a simple model of \gc{convergent adaptation in 
 a spatially spread population, where dispersal is local,
that is exposed to some novel, spatially homogeneous selection pressure.
We assumed that a single mutational change was sufficient to adapt
the population after the change in environment. Under this assumption,
selected alleles arise and spread locally.  See Figure
\ref{fig:sim-spread}. However, if the geographic
area is large enough multiple selected alleles can
arise independently and spread before any one has spread across all of
space. Therefore, one allele's rapid spatial spread can be halted by
running into convergent allele.A somewhat analogous situation also
arises in spatial models of clonal inference in asexuals \citep{gordo_adaptive_2006,martens_interfering_2011,otwinowski_clonal_2014}.}

@@

\gc{Under this model we previously derived} the characteristic geographic scale over which 
multiple instances of the adaptive allele are expected to arise in parallel. 
This characteristic length could be expressed in terms of a simple compound
parameter \mfp{comprised of} our parameters of interest. \gc{In
\citet{ralphcoop2010} we assumed that there was no standing variation for the adaptive
allele (e.g. because the allele was very strongly deleterious before
the environmental change)}, 
so that parallel mutation must be due to multiple new mutations occurring after the environmental shift.

@@

We assume that the species range
is a large, homogeneous, one-- or two--dimensional region. 
There are two selective classes -- the \gc{ {\em ancestral} } type,
and the {\em mutated} type \gc{(which will offer a fitness benefit after the
environmental switch)}.
We assume that separately arising mutations are distinguishable --
either as selectively equivalent mutations, or by linked \gc{neutral} variation.
\gc{We assume that all alleles of the mutated type are selectively
  equivalent (both in terms how deleterious they were before, and how
  advantagous they are after, the
  environmental shift). Note that these mutations do not have to arise at the same locus,
  just that all of the mutated type are selective equivalent. As such,
  these could be mutations arising at the same base pair or knockout
  mutations of any one of a number of genes in a pathway as long as carrying at least one of these alleles
  is sufficient to adapt an individual to the new environment.}

@@

For concreteness, suppose that before time $t=0$,
the mutant type has fitness $1-s_d$ relative to the neutral type
(i.e.\ it produces on average $1-s_d$ times the number of offspring per generation),
and that after time $t=0$,
the mutant type has fitness $1+s_b$,
where $s_b>0$ will usually be assumed to be small, and $0<s_d<1$.
We assume that diploid fitness is additive, or at least that the
important early dynamics are determined by the heterozygous fitness, with no reference to the fitness of the
homozygote for the mutant alleles. \mfp{Note that this means that we
  are ignoring the case of totally recessive alleles.}
\gc{We assume that the variance in offspring number is one, see
  \cite{ralphcoop2010} for how this assumption can be relaxed.}
%The number of offspring has finite variance, 
%which would divide the probability of establishment of advantageous alleles below,
%but to keep the formulas simpler we assume to be equal to one.
As for the other parameters,
suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
and that the mean squared geographic distance between parent and child
is $\sigma^2$. \mfp{We assume that the dispersal kernel is
  not heavy tailed, e.g. Gaussian, see \citet{ralphcoop2010} for
  discussion of heavy tailed kernels and \citet{hallatschek_acceleration_2014} for recent progress
  on rapid spreading due to very long distance dispersal.}
The species occupies an area with mean density
$\rho$ of alleles per unit area \gc{(twice the number of diploid
  individuals per unit area)}.

@@

\paragraph{Rates of origination of standing and new mutations}
We make use of the commonly used approximation that neglects competition between close relatives,
treating the offspring of a new mutant that appears in an area not already occupied by the mutated type
as a branching process \gc{in continuous time, measured in generations
(average inter-birth intervals)}. 
After $t=0$, the offspring of each individual mutant thus forms (approximately) a branching process with growth rate $s_b$,
so each new mutant establishes locally with probability $p_s \approx 2s_b$.
As in \cite{ralphcoop2010}, each new offspring has a very small probability of being a mutant and establishing locally,
so the collections of times and locations at which mutants appear and establish locally 
is well--approximated by a Poisson process in space and time.
\gc{By this we mean that there is a constant rate across space and
  time at which new
  mutations arise, and the occurance of new mutations in any region of
  space-time is independent of that in non-overlapping intervals.}
The rate of this Poisson process,
i.e.\ the mean number of new mutants per unit area and per generation that
appear and establish locally after $t=0$ in areas not already occupied by the mutant type,
is approximately 
\begin{equation}
\lambda = 2 s_b \mu \rho 
\end{equation}
\gc{In geographic areas already
 occupied by the mutant type (i.e. in regions were the beneficial
 allele has spread to), new mutations are selectively excluded,
 i.e. are effectively neutral, and so unlikely to establish over the
 short-time scales considered here.  }

Before $t=0$, on the other hand, the allele is deleterious.
The genetic descendants of each new mutation are (with high probability) doomed to extinction,
but may persist for \mfp{a short} time (note that we have assumed that $s_d$ is not too small).
\gc{The} times and locations of mutants before $t=0$ also \gc{will} be 
well--approximated by a 
Poisson process with rate $\mu \rho$ \gc{(if $\mu \rho \ll 1$) as
mutations are rare events. Therefore,} 
the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
by the Poisson Mapping Theorem, a Poisson process with rate we define
to be $\lambda_0$ \gc{(i.e. we can simply modify/thin the original rate)}.
\mfp{Our $\lambda_0$ is the mean density
of standing variants that are present and escape loss when the
environment shifts.}
% (In principle, more than one member of a mutant family may fix,
% but we assume these are close together enough that we may treat them as beginning from a single point.)
If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
and that these progenitors are near to each other in space, 
we can then treat each such family as equivalent to a single new mutation,
but with somewhat larger probability of local establishment,
\gc{following the environmental shift}.
(This approximation will be good 
if the logarithm of the size of each extant family is small relative to the establishment time,
and the spatial distribution of each is small relative to the spread between the families.)
To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let \gc{$Z_T$ be the number of its descendants at time $0$}.
At time $t=0$, when the environment shifts, there are $Z_T$ individuals present with the mutation,
and each has probability $p_s$ of establishing, approximately independently.
Therefore, the probability that at least one descendant of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
the mean number of clusters of standing variants destined to fix locally, per unit area, is
\begin{align*}
    \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
\end{align*}
For $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
resulting in the approximation
\begin{align}
    \lambda_0 &= \mu \rho \int_0^\infty 2s_b (1-s_d)^{T} dT \\
        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .\\
        & =  \frac{\lambda }{ (-log(1-s_d)) }.
\end{align}
Note that for small $s_d$ this can also be found by taking the expected frequency under mutation-selection
balance $\mu/s_d $ \citep{Haldane:27,Haldane:37} multiplying it by the
population density to obtain the expected number of chromosomes per unit
carrying the deleterious allele, with each of these having a
probability $2s_b$ of escaping loss \citep[an analogous approach to
that taken by ][]{Orr:01}
.

@@

If we ignore the effects of new mutation,
then everything about the process is relatively simple:
each point in space will be first reached by the wave whose origination point lies closest to it.
This random tessellation of space is known as a Poisson-Voronoi tessellation \citep{Moller:94}
(i.e.\ the cells formed by assigning regions of space to the nearest point in a Poisson process).
The properties of this tessellation by alleles is
determined by the spatial locations of the initiation points, 
which are sampled from a spatially homogeneous Poisson process, 
independent of the rate of spatial spread. 
Introducing new mutations cause some qualitative changes
beyond dependence on new parameters:
the cells formed by a Voronoi tessellation have straight sides,
but the introduction of new mutations cause these to curve \citep[see
Figure 1 of ][ for a graphical depiction of this point]{ralphcoop2010}.

\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=\textwidth]{spreading_alleles_trimmed_2}
  \end{center}
  \caption{
{\bf Cartoon space-time diagram of the geographic spread of standing and new mutations}. 
Time runs down the page, and a single spatial dimension runs across the page. 
The environment switches at $t=0$, 
following which adaptive alleles that successfully escape loss by drift
spread locally from both standing variants (lightning bolts) and new
mutations (stars). \gc{(Note that the standing variants arose at times
prior to $t=0$.)}
The mean spatial density of successful standing variants at $t=0$ is $\lambda_0$,
and the mean density in space $\times$ time of successful new mutations is $\lambda$.
These successful alleles spread spatially outwards at speed $v$, 
so the space-time profile of their spread forms a cone. Any mutation
that arises in an area already within a cone to will not have a selective advantage.
When these adaptive alleles meet, 
their rapid advance is halted because they no longer have an advantage,
so their boundaries form straight lines on short time-scales 
(over longer time-scales, they will mix into each other). 
\gc{One of the ways we derive various results in the text is by taking a
space-time point $(x,t)$ and asking the probability that no successful
alleles have spread there yet.} The grey dashed line shows a cone radiating backward in space-time,
originating $(x,t)$, with slope $v$, its radius at $t=0$ is $v t$. 
Our point $x$ at time $t$ is not yet adapted because no successful alleles appear within this grey cone.
}
  \label{fig:cartoon}
\end{figure}

@@

Over roughly the past ten thousand years, alleles conferring resistance to malaria have arisen in a number of genes 
and spread through human populations in areas where malaria \mfp{has been} endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to be examples of convergent adaptation, 
as different derived mutations in the same gene are seen in different individuals.
For example, 
a number of changes that confer malaria resistance have been observed in the $\beta$-globin gene;
and the sickle cell allele may plausibly have arisen by up to five independent occurrences
of the same base pair mutation at different locations within Africa
\citep{Flint:98,ralphcoop2010}.
Another particularly impressive case of convergent evolution \mfp{is} the numerous changes throughout the X-linked G6PD gene, 
with upward of 50 polymorphic variants (above $1\%$ local frequency) having so far been described 
that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. 
These alleles are now found at a combined frequency of around 8\% frequency in malaria endemic areas,
rarely exceeding 20\% \citep{Howes-g6pd-preval}. 
Whether these all confer resistance to malaria is unknown,
but malaria is thought to be the primary driver of these polymorphisms
\citep[see][ for a general review]{hedrick2011population}.
Three G6PD deficiency alleles are particularly common and relatively well studied: 
the $A-$ allele found in much of sub-Saharan Africa; 
the Med allele found in the Mediterranean and Middle East; 
and the Mahidol allele found Myanmar and Thailand.
\mfp{ 
The protective effects of these G6PD alleles are complicated, and are likely heterogeneous
across study populations and the form of malaria considered
\citep[see ][ for recent discussion]{manjurano_2015, malaria_network_2014}.  
The $A-$ and Mahidol alleles are thought to offer some protective effects against {\it Plasmodium falciparum} and {\it P.\ vivax} in both
heterozygote females and/or hemizygote males \citep{Ruwende-g6pd,
  Louicharoen-g6pd, manjurano_2015}. }
Haplotype-based analysis of genetic diversity surrounding
$A-$, Med, and Mahidol suggest that they have spread over the past 
few thousand years \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other known malaria resistance alleles. 
Population genetic analyses suggest that these three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$
\citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}. 
This is in reasonable agreement with the selection coefficients calculated by
\citet{Ruwende-g6pd} on the basis of the present day levels of
resistance to malaria due to the $A-$ allele. \\

@@

Given \mfp{such strong selection the alleles} should have risen
quickly to fixation, so their presence at intermediate frequency,
over a broad geographic area, makes it a good candidate for a recently
balanced polymorphism due to heterozygote advantage \citep[note that
the conditions for a balanced polymorphism are
complicated by the hemizygosity of males, see][]{hedrick2011population, Pamillo:1979}. 
Indeed, hemizygous males and homozygous females suffer from G6PD deficiency,
and homozygote females may also not be protected against malaria \citep{manjurano_2015, malaria_network_2014}. 
The theory we use regarding the ``wave of advance'' \citep{fisher1937wave}
applies as well in the case of heterozygote advantage
\citep{aronson1975nonlinear}, with the selected allele spreading
locally to the equilibrium frequency (rather than fixation). 
Therefore, our framework is applicable to the spread of G6PD, with speed determined by the advantage of heterozygotes when rare.
We assume that before malaria became prevalent, 
G6PD deficiency alleles suffered a decrease in relative fitness of $s_d$ in heterozygote and homozygotes females and hemizygote males. 
Assuming that the underlying causes and strength of this drop in fitness have not changed, 
we estimate that $s_d$ has to have been upward of $\sim 0.05$ (if $sb
\geq 0.05$),
in order to have resulted in the equilibrium frequency seen today in
areas with endemic malaria \citep[based on heterozygote advantage
calculations for the X chromosome, results not shown, see also][]{Ruwende-g6pd}.

@@

The geographic area of Central and Eastern Asia with malaria is 
on the order of ten million square kilometers.
In that area there are at least $15$ common variants \citep[see Figure \ref{fig-G6PD-map}][]{Howes-g6pd-variants}. 
Therefore, the average width of an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
The coding region of G6PD is 515 codons long, 
and around 140 distinct deficiency alleles have been observed. 
Assuming a mutation rate of $\approx 10^{-8}$ per base pair per generation, 
we can take as an order-of-magnitude estimate $\mu \approx 10^{-6}$ per generation. 
The dispersal and demographic parameters of humans in the past few thousand years is unclear,
particularly as we are concerned with the ``effective'' population density
(i.e.\ population density divided by variance in offspring number).
We therefore will use two reasonable values for the effective population density: $\rho=2$ and $0.2$ people per km$^2$,
and three values for the dispersal distance: $\sigma=10$, $50$ and $100$ kilometers per generation.
Clearly, human migration has been shaped both by local dispersal and larger-scale expansions 
\citep[see][for a recent discussion]{Pickrell:14}, 
so these parameters only provide a rough view of the process.

@@

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_chartimes_sd_sb.pdf}   %{charlen-by-sd-limit}
\caption{ %
{\bf Mean adaptation times} as a function of \gc{selective advantage and disadvantage 
as well as the population density.} We compare these to the corresponding quantities without standing variation and only considering standing variants.
} \label{G6PD_chartimes}
\end{center}
\end{figure}

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage[authoryear]{natbib}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{hyperref}
%\usepackage{latexml} % for \iflatexml
\usepackage{color}
\usepackage[utf8]{inputenc}  % for special characters in authors' names

\newcommand{\gc}[1]{{\it\color{blue} #1}}
\newcommand{\mfp}[1]{{\it\color{red} #1}}

@@

\section*{Abstract}
The extent to which populations experiencing shared selective pressures
adapt through a shared genetic response
is relevant to many questions in evolutionary biology. 
In a number of well studied traits and species, 
it appears that convergent evolution within species is common. 
In this paper, 
we explore how standing, genetic variation contributes to
convergent genetic responses in a geographically spread population,
extending our previous work on the topic.
Geographically limited dispersal slows the spread of each selected allele,
hence allowing other alleles 
-- newly arisen mutants or present as standing variation -- 
to spread before any one comes to dominate the population. 
When such alleles meet, their progress is substantially slowed 
-- if the alleles are selectively equivalent, 
they mix slowly, dividing the species range into a random tessellation,
which can be well understood by analogy to a Poisson process model of crystallization. 
In this framework, 
we derive the geographic scale over which a typical allele is expected to dominate, 
the time it takes the species to adapt as a whole,
and the proportion of adaptive alleles that arise from standing variation.
Finally, we explore how negative pleiotropic effects of alleles before an environment change 
can bias the subset of alleles that contribute to the species' adaptive response. 
We apply the results to the many geographically localized G6PD
deficiency alleles thought to confer resistance to malaria, \mfp{where the large
mutational target size makes it a likely candidate for adaptation
from standing variation, despite the selective cost of G6PD
deficiency alleles in the absence of malaria.}
We find the numbers and geographic spread of these alleles matches our predictions reasonably well,
\mfp{consistent with the view that they arose from a combination of} 
standing variation and new mutations since the advent of malaria.
Our results suggest that much of adaptation may be geographically local 
even when selection pressures are homogeneous. 
\mfp{Therefore, we argue that caution must be exercised when arguing that strongly
geographically restricted alleles are necessarily the outcome of local adaptation.}
We close by discussing the implications of these results for 
ideas of species coherence and the nature of divergence between species.

@@

Spatial population structure,
as caused for example by geographically limited dispersal, 
also increases the chance of convergent evolution. 
For example, 
\gc{geographically patchy} selection pressures can lead to much higher probability of parallel adaptation than uniform pressures,
since alleles are unable to spread through intervening populations
\citep{RalphCoop:14}. 
In \citet{ralphcoop2010} we formulated a simple model of \gc{convergent adaptation in 
a spatially spread population with local dispersal
that is exposed to some novel, spatially homogeneous selection pressure.
We assumed that a single mutational change was sufficient to adapt
the population after the change in environment. Under this assumption,
selected alleles arise and spread locally, as shown in Figure \ref{fig:sim-spread}. 
If the geographic area is large enough multiple selected alleles can
arise independently and spread before any one has spread across all of space. 
A somewhat analogous situation also
arises in spatial models of clonal inference in asexuals \citep{gordo_adaptive_2006,martens_interfering_2011,otwinowski_clonal_2014}.}

@@

\gc{Under this model we previously derived} the characteristic geographic scale over which 
multiple instances of the adaptive allele are expected to arise in parallel, 
a characteristic length \mfp{expressed in terms the parameters of interest.} 
\gc{In \citet{ralphcoop2010} we assumed that there was no standing variation for the adaptive
allele (e.g., because the allele was very strongly deleterious before the environmental change)}, 
so that parallel mutation must be due to multiple new mutations occurring after the environmental shift.

@@

We assume that the species range
is a large, homogeneous, one-- or two--dimensional region. 
There are two selective classes -- the \gc{\textit{ancestral}} type,
and the {\em mutated} type \gc{(which will offer a fitness benefit after the
environmental switch)}.
We assume that separately arising mutations are distinguishable --
either as selectively equivalent mutations, or by linked \gc{neutral} variation.
\gc{We assume that all alleles of the mutated type are selectively
equivalent (both in terms how deleterious they were before, and how
advantagous they are after, the environmental shift). 
Note that these mutations do not have to arise at the same locus,
just that they are selectively equivalent:
these could be mutations arising at the same base pair, 
or knockout mutations of any one of a number of genes in a pathway,
as long as carrying at least one of these alleles
is sufficient to adapt an individual to the new environment.}

@@

For concreteness, suppose that before time $t=0$,
the mutant type has fitness $1-s_d$ relative to the neutral type
(i.e.\ it produces on average $1-s_d$ times the number of offspring per generation),
and that after time $t=0$,
the mutant type has fitness $1+s_b$,
where $s_b>0$ will usually be assumed to be small, and $0<s_d<1$.
We assume that diploid fitness is additive, or at least that the
important early dynamics are determined by the heterozygous fitness, with no reference to the fitness of the
homozygote for the mutant alleles. \mfp{Note that this means that we
  are ignoring the case of totally recessive alleles.}
\gc{We assume that the variance in offspring number is one, see
  \cite{ralphcoop2010} for how this assumption can be relaxed.}
%The number of offspring has finite variance, 
%which would divide the probability of establishment of advantageous alleles below,
%but to keep the formulas simpler we assume to be equal to one.
As for the other parameters,
suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
and that the mean squared geographic distance between parent and child
is $\sigma^2$. \mfp{We assume that the dispersal kernel is
  not heavy tailed, i.e., falling into the Gaussian domain of attraction, see \citet{ralphcoop2010} for
  discussion of heavy tailed kernels and \citet{hallatschek_acceleration_2014} for recent progress
  on rapid spreading due to very long distance dispersal.}
The species occupies an area with mean density
$\rho$ of alleles per unit area \gc{(twice the number of diploid
  individuals per unit area)}.

@@

\paragraph{Rates of origination of standing and new mutations}
We make use of the commonly used approximation that neglects competition between close relatives,
treating the offspring of a new mutant that appears in an area not already occupied by the mutated type
as a branching process \gc{in continuous time, measured in generations
(average parental age at birth)}. 
After $t=0$, the offspring of each individual mutant thus forms (approximately) a branching process with growth rate $s_b$,
so each new mutant establishes locally with probability $p_s \approx 2s_b$.
As in \cite{ralphcoop2010}, each new offspring has a very small probability of being a mutant and establishing locally,
so the collections of times and locations at which mutants appear and establish locally 
is well--approximated by a Poisson process in space and time.
\gc{By this we mean that there is a constant rate across space and
  time at which new
  mutations arise, and the occurance of new mutations in any region of
  space-time is independent of that in non-overlapping intervals
  in which the mutated type has not already appeared.}
The rate of this Poisson process,
i.e.\ the mean number of new mutants per unit area and per generation that
appear and establish locally after $t=0$ in areas not already occupied by the mutant type,
\plr{is the product of the haploid population density, denoted $\rho$,
the mutation rate ($\mu$),
and the probability of fixation $p_s$, or}
approximately 
\begin{equation}
\lambda = 2 s_b \mu \rho 
\end{equation}
\gc{In geographic areas already occupied by the mutant type, 
 new mutations are effectively neutral, and so unlikely to establish over the
 short-time scales considered here, and are hence excluded.  }

Before $t=0$, on the other hand, the allele is deleterious.
The genetic descendants of each new mutation are (with high probability) doomed to extinction,
but may persist for \mfp{a short} time (note that we have assumed that $s_d$ is not too small).
\gc{The} times and locations of new mutations before $t=0$ also \gc{will} be 
well--approximated by a 
Poisson process with rate $\mu \rho$ % PLR: this is not required, and depends on the units you measure area in: (if $\mu \rho \ll 1$) 
\gc{since mutations are rare events. Therefore,} 
the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is 
a thinning of the original process, 
and so, by the Poisson Mapping Theorem, is also a Poisson process.
\mfp{We define $\lambda_0$ to be the mean density of this process, 
i.e., the geographic density
of standing variants that are present and escape loss when the
environment shifts.}
% (In principle, more than one member of a mutant family may fix,
% but we assume these are close together enough that we may treat them as beginning from a single point.)
If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
and that these progenitors are near to each other in space, 
we can then treat each such family as equivalent to a single new mutation,
but with somewhat larger probability of local establishment,
\gc{following the environmental shift}.
(This approximation will be good 
if the logarithm of the size of each extant family is small relative to the establishment time,
and the spatial distribution of each is small relative to the spread between the families.)
To find $\lambda_0$, consider a mutation that arose \plr{$T$ generations ago}, 
and let \gc{$Z_T$ be the number of its descendants at time $0$}.
At time $t=0$, when the environment shifts, there are $Z_T$ individuals present with the mutation,
and each has probability $p_s$ of establishing, approximately independently.
Therefore, the probability that at least one descendant of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
the mean number of clusters of standing variants destined to fix locally, per unit area, is
\begin{align*}
    \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
\end{align*}
For $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
resulting in the approximation
\begin{align}
    \lambda_0 &= \mu \rho \int_0^\infty 2s_b (1-s_d)^{T} dT \\
        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .\\
        & =  \frac{\lambda }{ (-\log(1-s_d)) }.
\end{align}
Note that for small $s_d$ this can also be found by taking the expected frequency under mutation-selection
balance $\mu/s_d $ \citep{Haldane:27,Haldane:37} multiplying it by the
population density to obtain the expected number of chromosomes per unit area
carrying the deleterious allele, with each of these having a
probability $2s_b$ of escaping loss \citep[an analogous approach to
that taken by ][]{Orr:01}
.

@@

If we ignore the effects of new mutation,
then everything about the process is relatively simple:
each point in space will be first reached by the wave whose origination point lies closest to it.
This random tessellation of space is known as a Poisson-Voronoi tessellation \citep{Moller:94}
(i.e.\ the cells formed by assigning regions of space to the nearest point in a Poisson process).
The properties of this tessellation by alleles is
determined by the spatial locations of the initiation points, 
which are sampled from a spatially homogeneous Poisson process, 
independent of the rate of spatial spread. 
Introducing new mutations cause some qualitative changes
beyond dependence on new parameters:
the cells formed by a Voronoi tessellation have straight sides,
but the introduction of new mutations cause these to curve.
% \citep[see Figure 1 of ][ for a graphical depiction of this point]{ralphcoop2010}.

\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=\textwidth]{spreading_alleles_trimmed_2}
  \end{center}
  \caption{
{\bf Cartoon space-time diagram of the geographic spread of standing and new mutations}. 
Time runs down the page, and a single spatial dimension runs across the page. 
The environment switches at $t=0$, 
following which adaptive alleles that successfully escape loss by drift
spread locally from both standing variants (lightning bolts) and new
mutations (stars). \gc{(Note that the standing variants arose at times
prior to $t=0$.)}
The mean spatial density of successful standing variants at $t=0$ is $\lambda_0$,
and the mean density in (space $\times$ time) of successful new mutations is $\lambda$.
These successful alleles spread spatially outwards at speed $v$, 
so the space-time profile of their spread forms a cone. Any mutation
that arises in an area already within a cone will not have a selective advantage.
When these adaptive alleles meet, 
their rapid advance is halted because they no longer have an advantage,
so their boundaries form straight lines on short time-scales 
(over longer time-scales, they will mix into each other). 
\gc{One of the ways we derive various results in the text is by taking a
space-time point $(x,t)$ and asking the probability that no successful
alleles have spread there yet.} The grey dashed line shows a cone radiating backward in space-time,
originating at $(x,t)$, with slope $v$; its radius at $t=0$ is $v t$. 
This shows the location $x$ is not yet adapted at time $t$ because no successful alleles appear within this grey cone.
}
  \label{fig:cartoon}
\end{figure}

@@

Over roughly the past ten thousand years, alleles conferring resistance to malaria have arisen in a number of genes 
and spread through human populations in areas where malaria \mfp{has been} endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to be examples of convergent adaptation, 
as different derived mutations in the same gene are seen in different individuals.
For example, 
a number of changes that confer malaria resistance have been observed in the $\beta$-globin gene;
and the sickle cell allele may plausibly have arisen by up to five independent occurrences
of the same base pair mutation at different locations within Africa
\citep{Flint:98,ralphcoop2010}.
Another particularly impressive case of convergent evolution \mfp{is presented by} the numerous changes throughout the X-linked G6PD gene, 
with upward of 50 polymorphic variants (above $1\%$ local frequency) having so far been described 
that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. 
These alleles are now found at a combined frequency of around 8\% frequency in malaria endemic areas,
rarely exceeding 20\% \citep{Howes-g6pd-preval}. 
Whether these all confer resistance to malaria is unknown,
but malaria is thought to be the primary driver of these polymorphisms
\citep[see][ for a general review]{hedrick2011population}.
Three G6PD deficiency alleles are particularly common and relatively well studied: 
the $A-$ allele found in much of sub-Saharan Africa; 
the Med allele found in the Mediterranean and Middle East; 
and the Mahidol allele found Myanmar and Thailand.
\mfp{ 
The protective effects of these G6PD alleles are complicated, and are likely heterogeneous
across study populations and the form of malaria considered
\citep[see ][ for recent discussion]{manjurano_2015, malaria_network_2014}.  
The $A-$ and Mahidol alleles are thought to offer some protective effects against {\it Plasmodium falciparum} and {\it P.\ vivax} in both
heterozygote females and/or hemizygote males \citep{Ruwende-g6pd,
  Louicharoen-g6pd, manjurano_2015}. }
Haplotype-based analysis of genetic diversity surrounding
$A-$, Med, and Mahidol suggest that they have spread over the past 
few thousand years \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other known malaria resistance alleles. 
Population genetic analyses suggest that these three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$
\citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}. 
This is in reasonable agreement with the selection coefficients calculated by
\citet{Ruwende-g6pd} on the basis of the present day levels of
resistance to malaria due to the $A-$ allele. \\

@@

Given \mfp{such strong selection the alleles} should have risen
quickly to fixation, so their presence at intermediate frequency,
over a broad geographic area, makes it a good candidate for a recently
balanced polymorphism due to heterozygote advantage \citep[note that
the conditions for a balanced polymorphism are
complicated by the hemizygosity of males, see][]{hedrick2011population, Pamillo:1979}. 
Indeed, hemizygous males and homozygous females suffer from G6PD deficiency,
and homozygote females may also not be protected against malaria \citep{manjurano_2015, malaria_network_2014}. 
The theory we use regarding the ``wave of advance'' \citep{fisher1937wave}
applies as well in the case of heterozygote advantage
\citep{aronson1975nonlinear}, with the selected allele spreading
locally to the equilibrium frequency (rather than fixation). 
Therefore, our framework is applicable to the spread of G6PD, with speed determined by the advantage of heterozygotes when rare.
We assume that before malaria became prevalent, 
G6PD deficiency alleles suffered a decrease in relative fitness of $s_d$ in heterozygote and homozygotes females and hemizygote males. 
Assuming that the underlying causes and strength of this drop in fitness have not changed, 
we estimate that $s_d$ has to have been upward of $\sim 0.05$ (if $s_b \geq 0.05$),
in order to have resulted in the equilibrium frequency seen today in
areas with endemic malaria \citep[based on heterozygote advantage
calculations for the X chromosome, results not shown, see also][]{Ruwende-g6pd}.

@@

The geographic area of Central and Eastern Asia with malaria is 
on the order of ten million square kilometers.
In that area there are at least $15$ common, \plr{clinically relevant} variants \citep[see Figure \ref{fig-G6PD-map}, from][]{Howes-g6pd-variants}. 
\plr{(These are type 2 variants that express at ${}<50\%$ enzyme activity, 
predispose individuals to haemolytic anaemia, 
and are found in at least 10 localities;
see \citet{Howes-g6pd-variants} for more details.)}
Therefore, the average width of an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
The coding region of G6PD is 515 codons long, 
and around 140 distinct deficiency alleles have been observed. 
Assuming a mutation rate of $\approx 10^{-8}$ per base pair per generation, 
we can take as an order-of-magnitude estimate $\mu \approx 10^{-6}$ per generation. 
The dispersal and demographic parameters of humans in the past few thousand years is unclear,
particularly as we are concerned with the ``effective'' population density
(i.e.\ population density divided by variance in offspring number).
We therefore will use two reasonable values for the effective population density: $\rho=2$ and $0.2$ people per km$^2$,
and three values for the dispersal distance: $\sigma=10$, $50$ and $100$ kilometers per generation.
Clearly, human migration has been shaped both by local dispersal and larger-scale expansions 
\citep[see][for a recent discussion]{Pickrell:14}, 
so these parameters only provide a rough view of the process.

@@

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_chartimes_sd_sb.pdf}   %{charlen-by-sd-limit}
\caption{ %
{\bf Mean adaptation times} as a function of \gc{selective advantage and disadvantage.}
We compare these to the corresponding quantities without standing variation and only considering standing variants.
} \label{G6PD_chartimes}
\end{center}
\end{figure}
